Data Specialization for a Staged Lambda Calculus

### 2015-07-01 paper/splitting.tex
working on splitting section

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$
which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.

First pass evaluation reduces $p$ to $b$,
which is plugged in for $l$ to produce $[b/l]r$.
Correctness of splitting states that this should be equivalent 
to the residual produced by direct evaluation, $q$.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).
More precisely, we can state correctness as \ref{thm:allCorrect}:
\begin{theorem}
\label{thm:allCorrect}
If $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$ and $\splitone e A p {l.r}$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}
where $\reduce e v$ indicates standard monostage reduction of the term $e$ to the value $v$,
and $e_1 \equiv e_2$ indicates a monostage equivalence (defined later).

We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
In the base case, where $e$ is already a residual of the form $\exv q$, we know
$\splitone {\exv q} A {\exv{\tup{}}} {\_.q}$, and so by uniqueness of splitting, 
$p = b = \tup{}$ and $r = q$.
It directly follows that $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\exv v$ (via repeated $\stepsym \bbonem$), 
splitting $e$ (via the new relation~$\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$. 
Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first-stage results
contained in $v$.  For example, in the simple case where $e$
contains entirely first-stage computation, such as:
\begin{lstlisting}
1` 3`gr{`1`literalone+literaltwo`3`}`
\end{lstlisting}
the first-stage result is the value 3.  The second element of the pair, $b$,
is the boundary value used as input to the resumer in second pass
evaluation as $[b/l]r \equiv q$, where $q$ is the same as
result produced by second-stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second-stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second-stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first- and second-stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:stagedsemantics}), this annotation is
particularly useful for splitting. Specifically, splitting $\pure e$
terms trivially yields a unit resumer.

>>>>

Our splitting algorithm is given by the judgements $\splittwo e A p l r$, $\splitone e A c {l.r}$, and $v \vsplito \mval i q$,
which apply to \bbtwo\ terms, \bbonem\ terms, and partial values, respectively.
Before discussing the meaning of these outputs, 
we state a few theorems regarding the totality and uniqueness of splitting:

\begin{theorem} \leavevmode
\label{thm:total} 
\begin{itemize} 
\item For term $e$, if $\typestwo e A$, then $\splittwo e A p l r$. 
\item For term $e$, if $\typesone e A$, then $\splitone e A p {l.r}$.
\item For partial value $v$, if $\typesone v A$, then $v \vsplito \mval i q$.
\end{itemize}
\end{theorem}

\begin{theorem} \leavevmode
\label{thm:unique}
\begin{itemize} 
\item If $\splittwo e A p l r$ and $\splittwo e A {p'} {l'} {r'}$, then $p = p'$, $l = l'$, and $r = r'$.
\item If $\splitone e A p {l.r}$ and $\splitone e A {p'} {l'.r'}$, then $p = p'$, $l = l'$, and $r = r'$.
\item If $v \vsplito \mval i q$ and $v \vsplito \mval {i'} {q'}$, then $i = i'$ and $q = q'$.
\end{itemize}
\end{theorem}

The proofs are by induction on the typing derivation
and simultaneous induction on the splitting derivations, respectively.

\subsection{Correctness}

For any $\coltwo e A$
which reduces to the residual $\exv q$ (via repeated $\stepsym \bbtwo$),
splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.

First pass evaluation reduces $p$ to $b$,
which is plugged in for $l$ to produce $[b/l]r$.
Correctness of splitting states that this should be equivalent 
to the residual produced by direct evaluation, $q$.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).
More precisely, we can state correctness as \ref{thm:allCorrect}:
\begin{theorem}
\label{thm:allCorrect}
If $\typestwo [\cdot] e A$, $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$,
then $\reduce p b$ and $[b/l]r \equiv q$.
\end{theorem}
In this theorem, $\reduce e v$ indicates standard monostage reduction of the term $e$ to the value $v$,
and $\equiv$ indicates a monostage equivalence, which we cover later.

We prove \ref{thm:allCorrect} by induction on the steps of evaluation.  
In the base case, where $e$ is already a residual of the form $\exv q$, we know
$\splittwo {\exv q} A {\exv{\tup{}}} \_ q$, and so by uniqueness of splitting, 
$p = b = \tup{}$ and $r = q$.
It directly follows that $\reduce {\exv {\tup {}}} {\tup {}}$ and $q \equiv q$.
In the recursive case, we have 
$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
To help this case, we introduce to a new lemma, whose proof will concern the rest of this section:
\begin{lemma}
\label{lem:twoCorrect}
If $e \stepsym \bbtwo e'$, $\splittwo e A p  l r$, $\splittwo {e'} A {p'} {l'} {r'}$, and $\reduce {p'} {b'}$,
then $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed

All that remains is to prove \ref{lem:twoCorrect}.  
We will do this by induction on the derivation of the step.
However, since second-stage steps can depend on first-stage steps,
we first cover the definition of splitting terms at \bbonem.

For any $\colmix e A$, 
splitting $e$ (via the new relation~$\splitonesym$) again 
yields the pair of monostage terms $\pipeM c l r$. 
Evaluating $c$ in the first pass produces the pair
$\tup{i,b}$, where the first element $i$ represents the first stage part of the results of $e$,
and $b$ is the boundary value.
We can now define an analogue of \ref{lem:twoCorrect} for steps at \bbonem:
\begin{lemma}
\label{lem:oneCorrect}
If $e \stepsym \bbonem e'$, $\splitone e A p {l.r}$, $\splitone {e'} A {p'} {l'.r'}$, and $\reduce {c'} {\tup{i,b'}}$,
then $\reduce c {\tup {i,b}}$ and $[b/l]r \equiv [b'/l']r'$.
\end{lemma}
Like with the \bbtwo\ version, this requires that the boundary-value-applied resumer be equivalent both sides of the step.
But additionally, it also requires that the $i$ be exactly identical on both sides of the evaluation.

...

For example, in the simple case where $e$
contains entirely first-stage computation, such as:
\begin{lstlisting}
1` 3`gr{`1`literalone+literaltwo`3`}`
\end{lstlisting}
the first-stage result is the value 3.  The second element of the pair, $b$,
is the boundary value used as input to the resumer in second pass
evaluation as $[b/l]r \equiv q$, where $q$ is the same as
result produced by second-stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second-stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second-stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first- and second-stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

Rather than attempt global optimization of the outputs of splitting,
we instead leverage the type system to indicate when a term contains
no second-stage computations by adding a third world \bbonep\ whose
terms are purely first-stage.  While not necessary for ordinary staged
term evaluation (\ref{sec:semantics}), this annotation is
particularly useful for splitting. Specifically, splitting $\pure e$
terms trivially yields a unit resumer.
